---
title: "Evidence for ecological niche partitioning among ribbon and spotted seals in the Bering Sea and implications for their resiliency to climate change"
date: today
author:
  - name: Josh M. London
    email: josh.london@noaa.gov
    orcid: 0000-0002-3647-5046
    affiliations:
      - ref: noaa-mml

affiliations:
  - id: noaa-mml
    name: Alaska Fisheries Science Center
    department: National Marine Fisheries Service
    address: 7600 Sand Point Way NE
    city: Seattle
    state: WA
abstract: > 
  Ecoclogial niche partitioning allows sympatric species to co-exist in space and
  time while reducing direct competition for resources. In deep-diving seals
  (_Phocidae_) such niche partitioning can be observed as temporal differentiation
  in overlap, two-dimensional separation in space, three-dimensional separation
  in the water column, and variability in isotopic space. Here, we focus on two
  species of phocid seals in Bering Sea of Alaska -- ribbon seals
  (_Histriophoca fasciata_) and spotted seals (_Phoca largha_) -- and evidence for
  niche partitioning from two decades of bio-logger deployments and sampling of
  stable isotopes from multiple tissue types with different intergration times.
  
execute:
  include: false
---

```{r}
library(RPostgres)
library(dplyr)
library(dbplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(sf)
library(ctmm)
```

# Introduction

# Methods

```{r}
locs_sf <- readRDS(here::here('data/locs_sf.rds'))
```

A total of `r locs_sf |> dplyr::distinct(ptt) |> nrow()` bio-loggers (SPLASH
family, Wildlife Computers, Redmond, Washington, USA) were deployed on
`r locs_sf |> dplyr::filter(species == 'Ribbon seal') |> distinct(speno) |> nrow()`
ribbon seals and 
`r locs_sf |> dplyr::filter(species == 'Spotted seal') |> distinct(speno) |> nrow()`
spotted seals between `r min(locs_sf$deploy_dt) |> lubridate::year()` and
`r max(locs_sf$end_dt) |> lubridate::year()`. The deployments span all age classes
with the exception of dependent pups for both species and were deployed during the
months of April, May, and June. In some cases, deployments were initiated prior to molting
and the bio-loggers fell off after a period of weeks to two months. Deployments
initiated after molting transmitted up to 9 months.

All deployments were checked for any data quality issues and inconsistent
location estimates before they were run through a course speed filter to
remove any locations that would have required a sustained swim speed greater
than 20 km/h. Additionally, any deployments with fewer than 30 location
estimates or a total deployment length less than 7 days were removed.

Two data sets for each species were created to separate movement in the months of April, May,
and June from the rest of the year. The continuous time movement model used in
the analysis is predicated on a general range limitation to the underlying
movement behavior. Both species have known association with the marginal
sea-ice zone during those months as they focus on pupping, breeding, and
molting. Initially, no seasonal divisions are specified for the remainder of
the year but may be required in subsequent analyses.

```{r}
filter_tracks <- function(tracks_sf) {
  
  crs <- sf::st_crs(tracks_sf)
  dat <- tracks_sf %>%
    sf::st_transform(4326) %>%
    ungroup() %>%
    arrange(deployid, locs_dt)
  dat_tr <- trip::trip(dat, c("locs_dt","deployid"), correct_all = FALSE)
  
  suppressWarnings(
    keep <- trip::sda(
      dat_tr,
      smax = 20 #km/hour
    )
  )
  
  tracks_filt <- dat %>%
    mutate(sda_keep = keep) %>%
    filter(sda_keep) %>%
    dplyr::select(-c(sda_keep, rank)) %>%
    st_transform(crs)
  return(tracks_filt)
}

locs_sf <- locs_sf |> 
  filter_tracks()

locs_sf_spring <- locs_sf |> 
  dplyr::filter(tag_type %in% c('SPLA','SGPS')) |> 
  dplyr::filter(lubridate::month(locs_dt) %in% c(3,4,5)) |> 
  dplyr::arrange(deployid,locs_dt) 

locs_sf_open_water <- locs_sf |> 
  dplyr::filter(tag_type %in% c('SPLA','SGPS')) |> 
  dplyr::filter(!lubridate::month(locs_dt) %in% c(3,4,5)) |> 
  dplyr::arrange(deployid,locs_dt) 

```

Utilization distributions were estimated for each species and each of the
seasonal periods based on a continuous time movement model (R package `ctmm`).
Specifically, optimally weighted auto-correlated kernel density estimates (wAKDE)
were created to reflect a more honest account of space use while also mitigating
sampling bias from irregular deployment lengths. The weighted AKDE utilization
distributions were combined into a population kernel density estimate that
should better reflect spatial distribution of the broader population beyond
just the sampled seals.

```{r}
as_telem <- function(locs_sf, out_proj) {
  locs_df <- locs_sf |> 
    sfheaders::sf_to_df(fill = TRUE)
  # separate fastloc and argos
  locs_f <- locs_df |> filter(type%in%c("FastGPS","known","User"))
  locs_a <- locs_df |> filter(type=="Argos")
  rm(locs_df)
  
  # rename for movebank conventions and convert
  locs_a <- locs_a |>
    rename(
      individual.local.identifier = deployid,
      timestamp = locs_dt,
      location.long = x,
      location.lat = y,
      Argos.orientation = error_ellipse_orientation,
      Argos.semi.minor = error_semi_minor_axis,
      Argos.semi.major = error_semi_major_axis
    ) %>% mutate(
      Argos.location.class = quality,
      quality = as.character(quality)
    )
  locs_a <- ctmm::as.telemetry(object = locs_a, projection = out_proj)
  locs_a <- tibble(deployid=names(locs_a), telem=locs_a)

  locs_f <- locs_f |>
    rename(
      individual.local.identifier = deployid,
      timestamp = locs_dt,
      location.long = x,
      location.lat = y
    ) %>% mutate(
      HDOP = dplyr::case_when(
        type == "known" ~ sqrt(2),
        type=="FastGPS" & quality=="4" ~ sqrt(2)*(1163)/20,
        type=="FastGPS" & quality=="5" ~ sqrt(2)*(169)/20,
        type=="FastGPS" & quality=="6" ~ sqrt(2)*(71)/20,
        type=="FastGPS" & quality=="7" ~ sqrt(2)*(43)/20,
        type=="FastGPS" & quality=="8" ~ sqrt(2)*(34)/20,
        type=="FastGPS" & quality=="9" ~ sqrt(2)*(28)/20,
        type=="FastGPS" & quality=="10" ~ sqrt(2)*(24)/20,
        type=="FastGPS" & quality=="11" ~ sqrt(2),
        TRUE ~ Inf
      ),
      quality = as.character(quality)
    )
  locs_f <- ctmm::as.telemetry(object = locs_f, projection = out_proj)
  uere(locs_f) <- 20
  locs_f <- tibble(deployid=names(locs_f), telem=locs_f)


  locs_df <- bind_rows(locs_a, locs_f) |> group_by(deployid) |> nest()
  locs_df <- locs_df |> rowwise() |> mutate(
    data = list(data$telem |> ctmm::tbind())
  )
  
  names(locs_df$data) <- locs_df$deployid

  return(locs_df)
}
```

```{r}
pl_telem_spring <- as_telem(locs_sf_spring |> 
                              dplyr::filter(species == 'Spotted seal'),
                            out_proj = 'epsg:3571'
                            )
pl_telem_open_water <- as_telem(locs_sf_open_water |> 
                                  dplyr::filter(species == 'Spotted seal'),
                                out_proj = 'epsg:3571'
                                )
hf_telem_spring <- as_telem(locs_sf_spring |> 
                              dplyr::filter(species == 'Ribbon seal'),
                            out_proj = 'epsg:3571'
                            )
hf_telem_open_water <- as_telem(locs_sf_open_water |> 
                                  dplyr::filter(species == 'Ribbon seal'),
                                out_proj = 'epsg:3571'
                                )
```

```{r}
hf_fits_spring <- vector("list", length = nrow(hf_telem_spring))

for(i in seq_along(hf_telem_spring$data)) {
  guess <- ctmm.guess(hf_telem_spring$data[[i]], interactive = FALSE)
  hf_fits_spring[[i]] <- ctmm.select(hf_telem_spring$data[[i]], guess)
}

names(hf_fits_spring) <- names(hf_telem_spring$data)

hf_akde_spring <- akde(hf_telem_spring$data,hf_fits_spring,weights=TRUE)
col <- color(hf_akde_spring, by='individual')

hf_pkde_spring <- pkde(hf_telem_spring$data, hf_akde_spring)
```

```{r}
hf_fits_open_water <- vector("list", length = nrow(hf_telem_open_water))

for(i in seq_along(hf_telem_open_water$data)) {
  guess <- ctmm.guess(hf_telem_open_water$data[[i]], interactive = FALSE)
  hf_fits_open_water[[i]] <- ctmm.select(hf_telem_open_water$data[[i]], guess)
}

names(hf_fits_open_water) <- names(hf_telem_open_water$data)

hf_akde_open_water <- akde(hf_telem_open_water$data,hf_fits_open_water,weights=TRUE)

hf_pkde_open_water <- pkde(hf_telem_open_water$data, hf_akde_open_water)
```


